Binning Strategies and Related Loss for Large Data

### 2015-05-06 Paper.Rnw
Euclidean

\begin{itemize}
\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a set of tiles centered at $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$. By displaying frequency information using shaded tiles instead of individual points there is a loss of information about the exact location of the points. %\hh{XXX the frequency information is not what causes the loss}
\ktm{
\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not mapped to a continuous shading scale. While shade can be \textit{rendered} continuously in HSV color space, thus representing frequency exactly, a human reader can not \textit{extract} this information at the same precision due to limitations of  human cognition.  %A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively 
In order to model these limitations we 
introduce a second stage of binning by using a discrete color scale for displaying binned frequencies, $b_C(c_\ell)$, $\ell \in \{1,\dots,\mathscr{L}\}$. %This loss in precision is then quantified as {\it Frequency loss}. %, thus losing precision in the display of the true frequencies.
}
\end{itemize}

@@

When the individual points of a scatterplot are collapsed to bin centers to be displayed as tiles in a binned scatterplot there is a loss of the location information.  This can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.\ the bin centers). Note that other distance metrics could be used, but the euclidean distance has a desirable interpretability in $\mathbb{R}^2$. The \textit{total spatial loss}, $L^S$, is defined as 
\begin{eqnarray}\label{totspatloss}
L^S = \sum_{i=1}^{n} L_i^S = \sum_{i=1}^{n} \sqrt{ \left(x_i-b_X(x_i)\right)^2 + \left(y_i-b_Y(y_i)\right)^2 }
\end{eqnarray} 
%
where $L_i^S$ is the loss in the  $i$th observation. 
%
Figure~\ref{fig:spatlossdemo2} visually displays the spatial loss for the data from  Table~\ref{DataReductionExample} as a result of standard rectangular binning. Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as black points and gray crosses, respectively. The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.

@@

\ktm{The spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded. If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.  However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard deviations. }

@@

% \item {\it Visual Loss}, $L^\text{V}$, is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile. This visual center will generally different from the numerical mean, causing some bias in the perception. Thus visual loss for a bin, $L_i^\text{N}$, is the euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$. The visual loss is the sum of these bin distances.

@@

% Numerical loss results from the binning process as points are reduced to the numerical center point. The loss for an individual point, $L_i^\text{N}$, is the euclidean distance between the $i^{th}$ data point $(x_i,y_i)$ and the corresponding bin numerical center $(\bar{x}, \bar{y})$. The overall numerical loss is thus the sum of numerical losses for individual points ($L^\text{N} = \sum_{i=1}^n L_i^\text{N}$).
% Visual loss is due to the distance between the numerical centers and the visual centers. The visual loss for each bin, $L_j^\text{V}$, is the euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$. The overall visual loss is the sum of these bin distances ($L^\text{V} = \sum_{j=1}^{N_x} L_j^\text{V}$). 

>>>>

\begin{itemize}
\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a set of tiles centered at $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$. By displaying frequency information using shaded tiles instead of individual points there is a loss of information about the exact location of the points. %\hh{XXX the frequency information is not what causes the loss}

\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not mapped to a continuous shading scale. While shade can be \textit{rendered} continuously in HSV color space, thus representing frequency exactly, a human reader can not \textit{extract} this information at the same precision due to limitations of  human cognition.  %A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively 
In order to model these limitations we 
introduce a second stage of binning by using a discrete color scale for displaying binned frequencies, $b_C(c_\ell)$, $\ell \in \{1,\dots,\mathscr{L}\}$. %This loss in precision is then quantified as {\it Frequency loss}. %, thus losing precision in the display of the true frequencies.

\end{itemize}

@@

When the individual points of a scatterplot are collapsed to bin centers to be displayed as tiles in a binned scatterplot there is a loss of the location information.  This can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.\ the bin centers). Note that other distance metrics could be used, but the Euclidean distance has a desirable interpretability in $\mathbb{R}^2$. The \textit{total spatial loss}, $L^S$, is defined as 
\begin{eqnarray}\label{totspatloss}
L^S = \sum_{i=1}^{n} L_i^S = \sum_{i=1}^{n} \sqrt{ \left(x_i-b_X(x_i)\right)^2 + \left(y_i-b_Y(y_i)\right)^2 }
\end{eqnarray} 
%
where $L_i^S$ is the loss in the  $i$th observation. 
%
Figure~\ref{fig:spatlossdemo2} visually displays the spatial loss for the data from  Table~\ref{DataReductionExample} as a result of standard rectangular binning. Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as black points and gray crosses, respectively. The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all segments represents the total spatial loss, $L^S$.

@@

\ktm{The spatial loss is a Euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded. If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.  However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard deviations. }

@@

% \item {\it Visual Loss}, $L^\text{V}$, is an implicit loss due to the rendering;  visually we perceive the center of a bin to be the center of the tile. This visual center will generally different from the numerical mean, causing some bias in the perception. Thus visual loss for a bin, $L_i^\text{N}$, is the Euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$. The visual loss is the sum of these bin distances.

@@

% Numerical loss results from the binning process as points are reduced to the numerical center point. The loss for an individual point, $L_i^\text{N}$, is the Euclidean distance between the $i^{th}$ data point $(x_i,y_i)$ and the corresponding bin numerical center $(\bar{x}, \bar{y})$. The overall numerical loss is thus the sum of numerical losses for individual points ($L^\text{N} = \sum_{i=1}^n L_i^\text{N}$).
% Visual loss is due to the distance between the numerical centers and the visual centers. The visual loss for each bin, $L_j^\text{V}$, is the Euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$. The overall visual loss is the sum of these bin distances ($L^\text{V} = \sum_{j=1}^{N_x} L_j^\text{V}$). 
